hatchR: A toolset to predict hatch and emergence phenology in wild fishes

### 2025-03-04 inst/manuscript/paper.Rmd
try to fix hanging HTML link in Case Study 1 (failed)
final edits to manuscript and then render

```{r, echo = TRUE, fig.cap="Output of hatchR function `plot_check_temp()`, which is used as a visual data check on the raw `year_sim` data set. Users can set custom thresholds for minimum and maximum temperatures (dashed lines)."}
plot_check_temp(data = year_sim,
                dates = date,
                temperature = temp,
                temp_min = 0, # temp_min and max lines are 
                              # user customizable
                temp_max = 25)
```

@@

We demonstrate how the `fit_model()` function may be used to create custom parameterizations for species beyond the Salmonids included in `model_table`. To showcase its utility, we provide parameterizations for three warm-water species: Smallmouth Bass (*Micropterus dolomieu*) [@webster1948], Channel Catfish (*Ictalurus punctatus*) [@small2001], and Lake Sturgeon (*Acipenser fulvescens*) [@smith2005]. These species were selected due to their common use in aquaculture and sport fisheries, illustrating the broad applicability of the effective values approach. For conciseness, we present parameterization for Smallmouth Bass here, while the full implementation details for all species are available in the [paper.Rmd](https://github.com/bmait101/hatchR/blob/master/inst/manuscript/paper_pdf/paper.Rmd) on the GitHub project repository.

To demonstrate parameterizations, we generate a simulated thermal regime featuring an ascending thermograph with a mean temperature of 16 °C (again, available in paper.Rmd). Using this dataset, we apply the parameterized models for each species to predict hatch timing and total developmental duration (Figure 3). This example highlights the flexibility of **hatchR** im accommodating diverse fish species and environmental conditions, making it a valuable tool for researchers and managers working outside of Salmonid system.

@@

These model expressions are then applied in `predict_phenology()` to estimate the days to hatch and development:

@@

To demonstrate this case study, we use the **hatchR** graphical user interface available at [https://elifelts.shinyapps.io/hatchR_demo/](https://elifelts.shinyapps.io/hatchR/){.uri}. Users begin by uploading their temperature dataset through the `Import Data` window, selecting their file, specifying the appropriate temperature and date columns, and providing the date format (e.g., year-month-day or day-month-year). For this example, we use the `crooked_river` dataset, included in the Shiny app as a demo dataset. It can also be accessed directly at <https://github.com/bmait101/hatchR/blob/master/extdata/crooked_river.csv/>.

@@

In this example, the model predicts that the last Bull Trout will emerge before the June 1st road work target date. This suggests that the Fisheries Biologist can confidently approve the work in the area without concern for sediment disturbance impacting fish developing in the gravel. This type of predictive modeling helps managers make informed, science-based decisions that balance conservation priorities with necessary land-use activities.

@@

To predict hatch timing across these sites, we first set up the necessary models and data, using Bull Trout parameterization. (For conciseness, we omit this set up here, but full details are available in the `paper.Rmd` file in the GitHub repository for **hatchR**.) We then apply `predict_phenology()` across all 139 sites, running predictions for three representative spawn dates (Early: September 1st, Peak: September 15th, Late: September 30th).

@@

To date, effective value models have primarily been used to predict phenology in wild environments using species-specific parameterizations [@adelfio2024; @austin2019]. However, these models fundamentally represent reaction norms, meaning that temperature-development relationships are influenced by genetic variation, gene-environment interactions, and phylogenetic differences [@west-eberhard2003].

For example, @sparks2017 found no significant differences in developmental rates between populations in their study but did observe family-level genetic × environment interactions under different thermal regimes. Similarly, when they reparameterized their models using western Alaskan Sockeye Salmon, they found slower developmental rates compared to populations from Canada [@beacham1990], consistent with cogradient variation [@conover2009; @sparks2022]. These findings highlight the importance of considering not only developmental time predictions but also how the underlying statistical relationships inform micro- and macro-evolutionary processes in fishes.

@@

The models within **hatchR** can be customized in multiple ways beyond the examples provided. While our current framework focuses on predicting hatch or emergence timing, it could be adapted to other key developmental milestones, such as early embryonic stages (*e.g.*, eye-up in salmonids; [@velsen1980]), initiation of pelagic-larval dispersal in marine fish, where temperature regulates the transition from demersal to pelagic behavior, or current-mediated dispersal in riverine species, where larvae begin downstream drift at temperature-dependent thresholds.

@@

**hatchR** is fully open source and reproducible. Source code and data can be found at <https://github.com/bmait101/hatchR/>. The latest version will be archived upon acceptance of the manuscript.

>>>>

```{r, echo = TRUE, fig.cap="Output of hatchR function `plot_check_temp()`, which is used as a visual data check on the raw `year_sim` data set. Users can set custom thresholds for minimum and maximum temperatures (dashed lines)."}
plot_check_temp(data = year_sim,
                dates = date,
                temperature = temp,
                temp_min = 0, # temp_min and max lines are 
                temp_max = 25) # user customizable
```

@@

We demonstrate how the `fit_model()` function may be used to create custom parameterizations for species beyond the Salmonids included in `model_table`. To showcase its utility, we provide parameterizations for three warm-water species: Smallmouth Bass (*Micropterus dolomieu*) [@webster1948], Channel Catfish (*Ictalurus punctatus*) [@small2001], and Lake Sturgeon (*Acipenser fulvescens*) [@smith2005]. These species were selected due to their common use in aquaculture and sport fisheries, illustrating the broad applicability of the effective values approach. For concision, we present parameterization for Smallmouth Bass here, while the full implementation details for all species are available in the `paper.Rmd` on the GitHub project repository.

To demonstrate parameterizations, we generate a simulated thermal regime featuring an ascending thermograph with a mean temperature of 16 °C (again, available in paper.Rmd). Using this dataset, we apply the parameterized models for each species to predict hatch timing and total developmental duration (Figure 3). This example highlights the flexibility of **hatchR** im accommodating diverse fish species and environmental conditions, making it a valuable tool for researchers and managers working outside of Salmonid systems.

@@

These model expressions are then implemented in `predict_phenology()` to estimate the days to hatch and development:

@@

To demonstrate this case study, we use the **hatchR** graphical user interface available at  <https://elifelts.shinyapps.io/hatchR_demo/>. Users begin by uploading their temperature dataset through the `Import Data` window, selecting their file, specifying the appropriate temperature and date columns, and providing the date format (e.g., year-month-day or day-month-year). For this example, we use the `crooked_river` dataset, included in the Shiny app as a demo dataset. It can also be accessed directly at [https://github.com/bmait101/hatchR/blob/master/extdata/crooked_river.csv/](#0){.uri}.

@@

In this example, the model predicts that the last Bull Trout will emerge before the June 1st road work target date. This suggests that the Fisheries Biologist can confidently approve the work in the area without concern for sediment disturbance impacting fish developing in the gravel. This type of predictive modeling helps managers make informed, science-based decisions that balance conservation priorities with land-use activities.

@@

To predict hatch timing across these sites, we first set up the necessary models and data, using Bull Trout parameterization (for concision, we omit this set up here, but full details are available in the `paper.Rmd` file in the GitHub repository or in the Predict fish phenology: nested article on **hatchR**'s website) We then apply `predict_phenology()` across all 139 sites, running predictions for three representative spawn dates (Early: September 1st, Peak: September 15th, Late: September 30th).

@@

To date, effective value models have primarily been used to predict phenology in wild environments using species-specific parameterizations [@adelfio2024; @austin2019]. However, these models fundamentally represent reaction norms, meaning that temperature-development relationships are influenced by local adaptation, gene-environment interactions, and phylogenetic differences [@west-eberhard2003].

For example, @sparks2017 found no significant differences in developmental rates between populations in their study but did observe family-level genetic × environment interactions under different thermal regimes. Similarly, when they reparameterized their models using western Alaskan Sockeye Salmon, they found slower developmental rates compared to populations from Canada [@beacham1990], consistent with cogradient variation [@conover2009; @sparks2022]. These findings highlight the importance of considering how developmental rates are keyedto specific environments but also how these underlying statistical relationships inform micro- and macro-evolutionary processes in fishes.

@@

The models within **hatchR** can be customized in multiple ways beyond the examples provided. While our current framework focuses on predicting hatch or emergence timing, it could be adapted to other key developmental milestones not reliant on exogenous feeding, such as early embryonic stages (*e.g.*, eye-up; [@velsen1980]), initiation or cessation of pelagic-larval dispersal, or current-mediated dispersal in riverine species, though not all cases may be as specifically tied to temperature as hatch and emergence.

@@

**hatchR** is fully open source and reproducible. Source code and data can be found at <https://github.com/bmait101/hatchR/>. The Rmardkdown document with all the code to reproduce the examples from this manuscript is available at <https://github.com/bmait101/hatchR/blob/master/inst/manuscript/paper.Rmd>. The The latest version will be archived upon acceptance of the manuscript.
